### 2017-02-08 Thesis/latex/paperC.tex
paperC compiled

In nonlinear dispersive evolution equations, the competing effects of nonlinearity and dispersion 
make a number of interesting phenomena possible. In the current work, the focus is on the numerical
approximation of traveling-wave solutions of such equations.
We describe our efforts to write a dedicated \textsf{Python}
code which is able to compute traveling-wave solutions of nonlinear
dispersive equations in a very general form.
%\begin{equation*}
%u_t + [f(u)]_{x} + \Lop u_x = 0,
%\end{equation*}
%%\vskip -0.05cm
%where $\Lop$ is a self-adjoint operator, and $f$ is a real-valued function with 
%$f(0) = 0$ and $f'(0) = 0$.

The \textsf{SpecTraVVave} code uses a continuation method coupled with a spectral projection
to compute approximations of steady symmetric solutions of this equation.
The code is used in a number of situations to gain an
understanding of traveling-wave solutions. The first case is the Whitham
equation, where numerical evidence points to the conclusion that the main 
bifurcation branch features three distinct points of interest, namely a turning point, 
a point of stability inversion, and a terminal point which corresponds to a cusped wave.

The second case is the so-called modified Benjamin--Ono equation where
the interaction of two solitary waves is investigated. It is found
that two solitary waves may interact in such a way
that the smaller wave is annihilated. The third case concerns
the Benjamin equation which features two competing dispersive operators.
In this case, it is found that bifurcation curves of periodic traveling-wave
solutions may cross and connect high up on the branch in the nonlinear regime.
%Include keywords and mathematical subject classification numbers as needed.

@@

The content of the paper is structured as follows. A mathematical description of the numerical method of 
\textsf{SpecTraVVave} is given in \autoref{sec:spectral}. \autoref{sec:experiments} presents results of different experiments carried out with the package. Concluding remarks are given in \autoref{sec:conclusion}. A method for finding initial guesses for the solver is described in \autoref{sec:stokes}. \autoref{sec:implementation} contains a schematic of program and a description of its workflow.

@@

Traveling wave solutions to the equation~\eqref{int_gen1}
are to be computed in the form of a linear combination of cosine functions of different
wave-numbers, i.e., in the space
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation}
\SN = \mathrm{span}_{\R}\setc{\cos(lx)}{0 \leq l \leq N-1 }
.
\end{equation}
%%%%%%%%%===========%%%%%%%%%%%%%%%
This is a subspace of $L^2(0,2\pi)$, and the collocation points 
$x_n = \pi \frac{2n -1}{2N}$ for $n = 1,\dots, N$ are used to discretize the domain.  
If the required full wavelength of solutions is to be $L \neq 2\pi$, 
one can use a scaling on the $x$-variable. 
Defining the new variable
\begin{equation*}
x' = \frac{L}{2\pi}x, 
\end{equation*}
yields collocation points $x_n'$ and wavenumbers $\kappa_l$ defined by
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation*}
x_n' = \frac{L}{2} \frac{2n -1}{2N}, \quad\quad \kappa_l = \frac{2\pi}{L}l. 
\end{equation*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
%%%%%%%%%===========%%%%%%%%%%%%%%%
%%%%%%%%%===========%%%%%%%%%%%%%%%
We are seeking a function $\phi_N \in \SN$ that satisfies the equations
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation}
-c\phi_N(x_n') + f(\phi_N)(x_n') + \Lop^N\phi_N(x_n') = 0, \label{int_gen-disc}
\end{equation}
%%%%%%%%%===========%%%%%%%%%%%%%%%
at the collocation points $x_n'$.
The operator $\Lop^N$ is the discrete form of the operator $\Lop$, 
and $\phi_N$ is the discrete cosine representation of $\phi$
which is given by
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
\phi_N(x') &= \sum_{l = 0}^{N-1} \omega(\kappa_l)\Phi_N(\kappa_l)\cos(\kappa_{l}x'),\\
\omega(\kappa_l) &= 
\begin{cases}
\sqrt{1/N}, \quad \kappa_l = 0, \\
\sqrt{2/N}, \quad \kappa_l > 0,
\end{cases}                  \\
\Phi_N(\kappa_l) &= \omega(\kappa_l) \sum_{n = 1}^{N} \phi_N(x'_n) \cos(\kappa_l x'_n),
\end{align*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
where $\kappa_l = 0,\frac{2\pi}{L},\ldots, \frac{2\pi}{L}(N-1)$ are the scaled wavenumbers,
and $\Phi_N(\cdot)$ are the discrete cosine coefficients.
As the equation~\eqref{int_gen-disc} is enforced at the collocation points $x_n'$, 
one may evaluate the term $\Lop^N\phi_N$ using the matrix $\Lop^N(i,j)$ 
defined by
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
\Lop^N\phi_N(x'_i) &= \sum_{j=1}^{N}\Lop^N(i,j) \phi_N(x'_j),\\
\Lop^N(i,j) &= \sum_{l=0}^{N-1} \omega^2(\kappa_l) \alpha(\kappa_l) \cos(\kappa_l x'_i) \cos(\kappa_l x'_j),
\end{align*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
where $\alpha(\cdot)$ is the Fourier multiplier function of the operator $\Lop$.

@@

\subsection{Convergence.}
In order to test the numerical implementation of the discretization,
the method is applied to a case where the solution is known.
One such case is the KdV equation
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
u_t + u_x + \frac{3}{2} u u_x + \frac{1}{6}u_{xxx} = 0, 
%\label{KdV_sol*}
\end{align*}
which has a known solution, given in the form
\begin{align*}
u_{\mathrm{exact}}(x,t) = a \sech^2\Big( {\textstyle\sqrt{\frac{3a}{4}}} (x-c t)\Big),\\
\end{align*}
with $c=1+a/2$.
%%%%%%%%%===========%%%%%%%%%%%%%%%
%
Using the boundary equation~\eqref{solitary}, \textsf{SpecTraVVave} is capable of computing 
approximations to solitary wave solutions of nonlinear wave equations. 
Solitary wave solutions are treated as traveling waves with sufficiently long wavelength 
that have the wave trough at zero.
In case of the KdV equation solitary wave solutions have exponential decay, 
and therefore, considering the symmetry of solitary solutions, 
the half-wavelength of $30$ is considered for the comparison. 
%\itodo[DM]{as mentioned before: should we call the (full-wavelength )/2 as half-wavelength OR just wavelength. 
%This may lead to confusions for reviewers.}
Approximation errors are summarized in \autoref{t2}.
%
\begin{table}  [ht]
\centering
\begin{tabular}{rrrr}
  \toprule
Nb.\ of grid points & $\log_{10}(\|u_{\mathrm{exact}} - u\|_{L^{\infty}})$ & $\log_{10}(\|u_{\mathrm{exact}} - u\|_{L^2})$ & Ratio of $L^2$-errors\\
\midrule
32 & $-1.389$  & $-2.092$ & \\

@@

%%%%%%%%%===========%%%%%%%%%%%%%%%
%\begin{figure}[ht]
%\centering
%\includegraphics[width=0.47\textwidth]{convergence}      
%\caption{\small Graph of error estimates given in \autoref{t2}.}
%\label{L2convergence}
%\end{figure} 
%%%%%%%%%===========%%%%%%%%%%%%%%%

@@

\subsubsection{Termination of the waveheight-velocity bifurcation curve of the Whitham equation.}
\label{sec-W-term}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The waveheight-velocity bifurcation curve of the Whitham equation 
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align}
	u_t + \frac{3}{2} u u_x + Ku_x = 0, \qquad 
	\widehat{Ku}(k) = \sqrt{\frac{\tanh(k)}{k}},  \label{W}
\end{align}
%%%%%%%%%===========%%%%%%%%%%%%%%%
was studied numerically in~\cite{Ehrnstrom2013}. 	
An attempt was made to identify the termination point of the Whitham bifurcation curve. 
The investigation was limited by computational tools and complete results were not obtained. 
In particular, the authors could not confirm that traveling wave solutions do not exist 
past the point where the authors, based on pioneering work of Whitham
\cite{Whitham1967j} suspected a cusped solution.
In this section a number of numerical results on nature of the bifurcation curve
for the Whitham equation are presented.
Solutions to the equation~\eqref{W}
are computed in the form of traveling waves $u(x,t) = \phi(x - c t)$
and the homogeneous $(B = 0)$ integrated version the equation is considered:
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align}
-c \phi + \frac{3}{4} \phi^2 + K \phi = 0. \label{W-int}
\end{align}
%%%%%%%%%===========%%%%%%%%%%%%%%%
Special attention is given to relation between stability of solutions 
and their waveheight and velocity parameters, i.e., their position on the bifurcation curve. 
The following questions are under study: 
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{enumerate}[label=\alph*)] % chktex 9 chktex 10
\item Where does the bifurcation curve terminate?

\item Where on the bifurcation curve do solutions change their stability?

\item Is there any role that the turning point on the bifurcation curve plays?
\end{enumerate}
%%%%%%%%%===========%%%%%%%%%%%%%%%

@@

From the data given in the \autoref{t1}, one can deduce that for solutions with minimum speed 
and maximum $L^2$-norm exponential fit is better than polynomial. 
That is not the case for the terminal solution. 
Thus, the first two solutions are smooth and the terminal solution is nonsmooth. 
In fact, the polynomial fit is better than exponential for solutions that 
are between the maximum $L^2$-norm solution and the terminal solution.

>>>>

%In nonlinear dispersive evolution equations, the competing effects of nonlinearity and dispersion 
%make a number of interesting phenomena possible. In the current work, the focus is on the numerical
%approximation of traveling-wave solutions of such equations.
%We describe our efforts to write a dedicated \textsf{Python}
%code which is able to compute traveling-wave solutions of nonlinear
%dispersive equations in a very general form.
%%\begin{equation*}
%%u_t + [f(u)]_{x} + \Lop u_x = 0,
%%\end{equation*}
%%%\vskip -0.05cm
%%where $\Lop$ is a self-adjoint operator, and $f$ is a real-valued function with 
%%$f(0) = 0$ and $f'(0) = 0$.
%
%The \textsf{SpecTraVVave} code uses a continuation method coupled with a spectral projection
%to compute approximations of steady symmetric solutions of this equation.
%The code is used in a number of situations to gain an
%understanding of traveling-wave solutions. The first case is the Whitham
%equation, where numerical evidence points to the conclusion that the main 
%bifurcation branch features three distinct points of interest, namely a turning point, 
%a point of stability inversion, and a terminal point which corresponds to a cusped wave.
%
%The second case is the so-called modified Benjamin--Ono equation where
%the interaction of two solitary waves is investigated. It is found
%that two solitary waves may interact in such a way
%that the smaller wave is annihilated. The third case concerns
%the Benjamin equation which features two competing dispersive operators.
%In this case, it is found that bifurcation curves of periodic traveling-wave
%solutions may cross and connect high up on the branch in the nonlinear regime.
%%Include keywords and mathematical subject classification numbers as needed.

@@

The content of the paper is structured as follows. A mathematical description of the numerical method of 
\textsf{SpecTraVVave} is given in section \ref{sec:spectral}. Section \ref{sec:experiments} presents results of different experiments carried out with the package. Concluding remarks are given in section \ref{sec:conclusion}. 
%A method for finding initial guesses for the solver is described in \ref{sec:stokes}. \ref{sec:implementation} contains a schematic of program and a description of its workflow.

@@

Traveling wave solutions to the equation~\eqref{int_gen1}
are to be computed in the form of a linear combination of cosine functions of different
wave-numbers, i.e., in the space
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation}
\SN = \mbox{span}_{\R}\{ cos(lx) | 0 \leq l \leq N-1 \}
.
\end{equation}
%%%%%%%%%===========%%%%%%%%%%%%%%%
This is a subspace of $L^2(0,2\pi)$, and the collocation points 
$x_n = \pi \frac{2n -1}{2N}$ for $n = 1,\dots, N$ are used to discretize the domain.  
If the required full wavelength of solutions is to be $L \neq 2\pi$, 
one can use a scaling on the $x$-variable. 
Defining the new variable
\begin{equation*}
x' = \frac{L}{2\pi}x, 
\end{equation*}
yields collocation points $x_n'$ and wavenumbers $\kappa_l$ defined by
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation*}
x_n' = \frac{L}{2} \frac{2n -1}{2N}, \quad\quad \kappa_l = \frac{2\pi}{L}l. 
\end{equation*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
%%%%%%%%%===========%%%%%%%%%%%%%%%
%%%%%%%%%===========%%%%%%%%%%%%%%%
We are seeking a function $\phi_N \in \SN$ that satisfies the equations
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{equation}
-c\phi_N(x_n') + f(\phi_N)(x_n') + \Lop^N\phi_N(x_n') = 0, \label{int_gen-disc}
\end{equation}
%%%%%%%%%===========%%%%%%%%%%%%%%%
at the collocation points $x_n'$.
The operator $\Lop^N$ is the discrete form of the operator $\Lop$, 
and $\phi_N$ is the discrete cosine representation of $\phi$
which is given by
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
\phi_N(x') &= \sum_{l = 0}^{N-1} \omega(\kappa_l)\Phi_N(\kappa_l)\cos(\kappa_{l}x'),\\
\omega(\kappa_l) &= 
\begin{cases}
\sqrt{1/N}, \quad \kappa_l = 0, \\
\sqrt{2/N}, \quad \kappa_l > 0,
\end{cases}                  \\
\Phi_N(\kappa_l) &= \omega(\kappa_l) \sum_{n = 1}^{N} \phi_N(x'_n) \cos(\kappa_l x'_n),
\end{align*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
where $\kappa_l = 0,\frac{2\pi}{L},\ldots, \frac{2\pi}{L}(N-1)$ are the scaled wavenumbers,
and $\Phi_N(\cdot)$ are the discrete cosine coefficients.
As the equation~\eqref{int_gen-disc} is enforced at the collocation points $x_n'$, 
one may evaluate the term $\Lop^N\phi_N$ using the matrix $\Lop^N(i,j)$ 
defined by
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
\Lop^N\phi_N(x'_i) &= \sum_{j=1}^{N}\Lop^N(i,j) \phi_N(x'_j),\\
\Lop^N(i,j) &= \sum_{l=0}^{N-1} \omega^2(\kappa_l) \alpha(\kappa_l) \cos(\kappa_l x'_i) \cos(\kappa_l x'_j),
\end{align*}
%%%%%%%%%===========%%%%%%%%%%%%%%%
where $\alpha(\cdot)$ is the Fourier multiplier function of the operator $\Lop$.

@@

\subsection{Convergence.}
In order to test the numerical implementation of the discretization,
the method is applied to a case where the solution is known.
One such case is the KdV equation
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align*}
u_t + u_x + \frac{3}{2} u u_x + \frac{1}{6}u_{xxx} = 0, 
%\label{KdV_sol*}
\end{align*}
which has a known solution, given in the form
\begin{align*}
u_{\mathrm{exact}}(x,t) = a \sech^2\Big( {\textstyle\sqrt{\frac{3a}{4}}} (x-c t)\Big),\\
\end{align*}
with $c=1+a/2$.
%%%%%%%%%===========%%%%%%%%%%%%%%%
%
Using the boundary equation~\eqref{solitary}, \textsf{SpecTraVVave} is capable of computing 
approximations to solitary wave solutions of nonlinear wave equations. 
Solitary wave solutions are treated as traveling waves with sufficiently long wavelength 
that have the wave trough at zero.
In case of the KdV equation solitary wave solutions have exponential decay, 
and therefore, considering the symmetry of solitary solutions, 
the half-wavelength of $30$ is considered for the comparison. 
%\itodo[DM]{as mentioned before: should we call the (full-wavelength )/2 as half-wavelength OR just wavelength. 
%This may lead to confusions for reviewers.}
Approximation errors are summarized in \ref{t2}.
%
\begin{table}  [ht]
\centering
\begin{tabular}{rrrr}
  \toprule
Nb.\ of grid points & $\log_{10}(\|u_{\mathrm{exact}} - u\|_{L^{\infty}})$ & $\log_{10}(\|u_{\mathrm{exact}} - u\|_{L^2})$ & Ratio of $L^2$-errors\\
\midrule
32 & $-1.389$  & $-2.092$ & \\

@@

%%%%%%%%%===========%%%%%%%%%%%%%%%
%\begin{figure}[ht]
%\centering
%\includegraphics[width=0.47\textwidth]{convergence}      
%\caption{\small Graph of error estimates given in \autoref{Ben_disp_rel}ref{t2}.}
%\label{L2convergence}
%\end{figure} 
%%%%%%%%%===========%%%%%%%%%%%%%%%

@@

\subsubsection{Termination of the waveheight-velocity bifurcation curve of the Whitham equation.}
\label{sec-W-term}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The waveheight-velocity bifurcation curve of the Whitham equation 
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align}
	u_t + \frac{3}{2} u u_x + Ku_x = 0, \qquad 
	\widehat{Ku}(k) = \sqrt{\frac{\tanh(k)}{k}},  \label{W}
\end{align}
%%%%%%%%%===========%%%%%%%%%%%%%%%
was studied numerically in~\cite{Ehrnstrom2013}. 	
An attempt was made to identify the termination point of the Whitham bifurcation curve. 
The investigation was limited by computational tools and complete results were not obtained. 
In particular, the authors could not confirm that traveling wave solutions do not exist 
past the point where the authors, based on pioneering work of Whitham
\cite{Whitham1967j} suspected a cusped solution.
In this section a number of numerical results on nature of the bifurcation curve
for the Whitham equation are presented.
Solutions to the equation~\eqref{W}
are computed in the form of traveling waves $u(x,t) = \phi(x - c t)$
and the homogeneous $(B = 0)$ integrated version the equation is considered:
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{align}
-c \phi + \frac{3}{4} \phi^2 + K \phi = 0. \label{W-int}
\end{align}
%%%%%%%%%===========%%%%%%%%%%%%%%%
Special attention is given to relation between stability of solutions 
and their waveheight and velocity parameters, i.e., their position on the bifurcation curve. 
The following questions are under study: 
%%%%%%%%%===========%%%%%%%%%%%%%%%
\begin{itemize} % chktex 9 chktex 10
\item[a)] Where does the bifurcation curve terminate?

\item[b)] Where on the bifurcation curve do solutions change their stability?

\item[c)] Is there any role that the turning point on the bifurcation curve plays?
\end{itemize}
%%%%%%%%%===========%%%%%%%%%%%%%%%

@@

From the data given in the \ref{t1}, one can deduce that for solutions with minimum speed 
and maximum $L^2$-norm exponential fit is better than polynomial. 
That is not the case for the terminal solution. 
Thus, the first two solutions are smooth and the terminal solution is nonsmooth. 
In fact, the polynomial fit is better than exponential for solutions that 
are between the maximum $L^2$-norm solution and the terminal solution.
